Sensor Clusterization in D-optimal Design in Infinite Dimensional Linear Bayesian Inverse Problems

### 2020-10-03 new_paper.tex
Need to revise the introduction, write clear contribution and weaknesses and decide what to do with sylvester's determinant lemma

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note the clusterization that occurs
    when 6 sensors are used --- only 4 measurement locations are
    included. Observed data does not influence the pointwise
    variances, hence it is omitted.}
\end{figure}\label{fig:clusterization illustration}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with homogeneous Dirichlet boundary
conditions.

@@

\subsection{Bayesian Linear Inverse Problems}\label{section:abstract OED}
Let $\hil$ be a Hilbert space and $\ban$ a Banach space, both
infinite-dimensional. Consider $\fwd: \hil \to \ban$, a linear
operator (the ``forward operator''). We are interested in forward
operators that are strongly smoothing (have fast decaying singular
values --- the heat operator from section \ref{subsec:example} is a
prime example). We want to infer $\param \in \hil$, some parameter of
the dynamics given noisy observations of $\fwd \param$. We take a zero
mean Gaussian prior $\pr$ for $\param$ (otherwise the mean can just be
subtracted everywhere). Hence, $\param \sim \pr = \normal(0,\prcov)$
with some appropriate covariance operator $\prcov$
\cite{Stuart10}. Observations (measurements) are taken via the linear
observation operator $\obs \in ( \ban^* )^m$, where $m$ is the number
of observations allowed (e.g. sensors at our disposal). In an analogy
with linear algebra (where row vectors are thought of as linear
functionals), we think of $\obs$ as having ``rows'' $\meas_j,
j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \ban^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

@@

@@

\subsection{Overview}
From \eqref{eq:objective} we wish to characterize solution(s) of the
following optimization problem for $\tar$. %%: (\ban^*)^m \to \R$:
\begin{align}\label{eq:optimization}
  \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
  = \argmax_{\obs} \frac12 \log \det 
  (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
\end{align}
where $\obs$ is constrained to some allowed set of measurements
(e.g.\ pointwise evaluations on $[0,\pi]$).
%%, for the example in section \ref{subsec:example}).
We seek a formulation of the optimal design using Lagrange
multipliers. In section \ref{section:objective} we find the gradient
of $\tar$ and record it in \eqref{eq:tar grad}. In section
\ref{subsec:unit length} we relax the constraints on $\obs$ so that
the optimization problem in \eqref{eq:optimization} is analytically
tractable.  We find gradients for the new constraints and record them
in \eqref{eq:grad constraints}. Finally, in section
\ref{subsec:necessary} we use these to formulate the optimal design
problem as a Lagrange multipliers problem in equation
\eqref{eq:conditions}.

@@

Thus, if we take $\func_j^{*} = \meas_j$ and $C := (\fwd \prcov
\fwd^*)^{-1}$ in the above lemma, we conclude from the eigenproblem
\eqref{eq:mod conditions} that $\obs^*\obs$ has the same eigenvectors
as $\fwd \prcov \fwd^*$, namely $\{\ev_i\}_{i=1}^{\infty}$ (all but
finitely many of them with zero eigenvalue, since $\obs^*\obs$ is
finite-rank). Denote the corresponding non-zero eigenvalues of
$\obs^*\obs$ by $\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i
\geq k+1$. Using Sylvester's determinant theorem we get:

@@

\pgfplotstableread{
  Label     prior  optimal  sub-optimal 
  1         0.2    1.3           1.4
  2         0.8    0.7           0.3
  3         2.2    0             0.2
  4         3.5    0             0.1
}\optimalvsnot

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Characterizing an optimal design, compared to a
    sub-optimal design. Both designs are allowed two measurements,
    which means the blue area has accumulated height of $2$ in both
    plots. An optimal design (left) increases precision (reduces
    uncertainty) where it is lowest. A sub-optimal design (right) does
    not. In this example, a measurement is taken partly in the direction of the
    fourth eigenvector, where precision is highest.}
  \label{fig:optimal vs not}
\end{figure}

\pgfplotstableread{
  Label    prior  $o_1$  $o_2$   $o_3$   $o_4$   $o_5$         $o_6$    topper
  1         0.2    0.8    0.5     0.5    0.4    0.33333333  0.3333333  0.001
  2         0.8    0.2    0.5     0.5    0.4    0.33333333  0.3333333  0.001
  3         2.2    0      0       0      0.2    0.33333333  0.3333333  0.001
  4         3.5    0      0       0      0      0           0          0.001
}\testdata

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\testdata}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector,
      ]

      \addplot [fill=green!80]  table [y=prior, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=blue!60]   table [y=$o_1$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=red!60]    table [y=$o_2$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=black!60]  table [y=$o_3$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=orange!60] table [y=$o_4$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=cyan!60]   table [y=$o_5$, meta=Label, x expr=\coordindex] {\testdata};
      \addplot [fill=purple!60] table [y=$o_6$, meta=Label, x expr=\coordindex] {\testdata};

      \addlegendentry{prior}
      \addlegendentry{$o_1$}
      \addlegendentry{$o_2$}
      \addlegendentry{$o_3$}
      \addlegendentry{$o_4$}
      \addlegendentry{$o_5$}
      \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Posterior precision per eigenvector, after taking a
    measurement $\meas_i$. Each $\meas_i$ increases precision
    uniformly across the lowest precision eigenvectors. We see that
    $\meas_2$ and $\meas_3$ are repeated measurements, as well as
    $\meas_5$ and $\meas_6$.}
  \label{fig:clusterization}
\end{figure}

The argument above has the important consequence (illustrated in
figure \ref{fig:optimal vs not}) that the eigenvalues in the
eigenvalue problem \eqref{eq:eigenproblem} are all equal! This makes
perfect sense. First, we know the Lagrange multipliers tell us how
much we gain by relaxing the constraints and there is no reason for
any measurement to give us more than any other (otherwise we would put
more weight on the better measurement, as can be understood from the
discussion in the following paragraph). This result of the analysis
above is summarized in the theorem below.

@@

\subsection{Sequential Design}
The way clusterization can occur is illustrated in figure
\ref{fig:clusterization} and understood using theorem
\ref{thm:char}. The first measurement $\meas_1$ is optimal, since
$\eta_i, i=1,2$ are chosen to increase and equalize
$\sigma^2\lambda_i^{-1} +\eta_i, i=1,2$, which are lowest. In this
case, $\meas_1$ has no component in the direction of $\ev_i, i\geq
3$. We see that each measurement increases precision (reduces
uncertainty) only for the eigenvectors of lowest precision. Consider a
sequential optimization scenario. We first seek a measurement to
maximize our our utility, given the prior. From , the posterior
covariance does not depend on data observed, so we know it after
deciding on the first measurement.

the best
first measuremen, thIf the precision in these is lower than the
precision for the next eigenvector, a repeated measurement is optimal.

Results of section \ref{subsec:characterization} specifically, the
discussion at the end of said section, along with figure
\ref{figure:clusterization} and equation \eqref{eq:true target} show
how clusterization arises when an optimal design is saught.

@@

\subsection{Derivation}
We rely on corollary \ref{cor:same meas} below, which is proved in the
appendix. Denote $\obs = (\meas_1,\dots,\meas_m)^t$ and $\obsm :=
(\meas_1,\dots,\meas_{m-1})^t$.
\begin{restatable*}{corollary}{samemeas}\label{cor:same meas}
  If $\meas_m = \meas_j$ for some $1 \leq j \leq m-1$, then
  \begin{equation*}
    \tar(\obs) - \tar(\obsm) =
    \log \left ( 1 + \frac{\sigma^2
      \langle \fwd \postcovm \fwd^* \obsm^* \Sigmam^{-1} e_j,
      \obsm^* \Sigmam^{-1}e_j \rangle
    }{
      2 - \sigma^2 e_j^t\Sigmam^{-1}e_j 
    }       
    \right ).
  \end{equation*}
\end{restatable*}
From Corollary \ref{cor:same meas} we can deduce that

$$
\lim_{\sigma^2 \to 0} \tar(\obs) -\tar(\obsm) = 0
$$
Hence no increase in the design criterion is
achieved. This is clearly sub-optimal and $\obsm$ cannot be an optimal
design. We may conclude that for small measurement error levels, the
clusterization effect will be mitigated by the presence of a non-zero
model error. Since the design criterion is not defined for $\sigma^2 =
0$ and identical measurements, we cannot make a statement regarding
$\sigma^2 = 0$, except in the limiting sense described above.

\section{Conclusion}\label{section:conclusion}
Our relaxed model gives us insight to D-optimal designs. We see that
in our setting, uncertainty is not reduced in any direction
(eigenvector) before it is the largest uncertainty present. We show
how and why repeated measurements can give rise to a D-optimal
design. There is more work to be done in understanding exactly what
causes such designs to be strictly better than others.

@@

\begin{lemma}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M =
  m$, $m > k$. Then we can find $\func_j \in \R^k,j=1,\dots,m$ with
  $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that $AA^t =
  M$.
\end{lemma}
\begin{proof}
  Let us diagonalize $M$, so that $M = U D U^t$ with $D =
  \diag(d_1,\dots,d_k)$ and $U \in \R^{k \times k }$ orthogonal. Let $S
  \in \R^{k \times m}$ with $S_{ii} = \sqrt{d_{i}}$ and zeros
  otherwise. Define $A:= U S V^t$, where $V \in \R^{m \times m}$ is
  orthogonal and will be further restricted later. Then $AA^t = U
  SV^tVS^t U^t = UDU^t$, so $AA^t$ has the required eigenvalues and
  eigenvectors by construction. If we can choose $V$ such that $A$
  also satisfies the unit norm constraints we are done. These
  constraints are, for $j=1,\dots,m$:
  \begin{equation}\label{eq:V constraints}
   1 = [A^tA]_{jj} = [V S^tS V^t]_{jj},
  \end{equation}
  and we can expect to do this since we assumed $\tr{D} = m$.

>>>>

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note the clusterization that occurs
    when 6 sensors are used --- only 4 measurement locations are
    included. Observed data does not influence the pointwise
    variances, hence it is omitted.}
  \label{fig:clusterization illustration}
\end{figure}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with homogeneous Dirichlet boundary
conditions.

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hil$ be a Hilbert space and $\ban$ a Banach space, both
infinite-dimensional. Consider $\fwd: \hil \to \ban$, a linear
operator (the ``forward operator''). We are interested in forward
operators that are strongly smoothing (have fast decaying singular
values --- the heat operator from section \ref{subsec:example} is a
prime example). We want to infer $\param \in \hil$, some parameter of
the dynamics given noisy observations of $\fwd \param$. We take a zero
mean Gaussian prior $\pr$ for $\param$ (otherwise the mean can just be
subtracted everywhere). Hence, $\param \sim \pr = \normal(0,\prcov)$
with some appropriate covariance operator $\prcov$
\cite{Stuart10}. Observations (measurements) are taken via the linear
observation operator $\obs \in ( \ban^* )^m$, where $m$ is the number
of observations allowed (e.g. sensors at our disposal). In an analogy
with linear algebra (where row vectors are thought of as linear
functionals), we think of $\obs$ as having ``rows'' $\meas_j,
j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \ban^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

@@

\subsection{Sequential vs Simultaneous Optimization}\label{subsec:seq vs sim}
From \eqref{eq:objective} we wish to characterize solution(s) of the
following optimization problem for $\tar$. %%: (\ban^*)^m \to \R$:
\begin{align}\label{eq:optimization}
  \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
  = \argmax_{\obs} \frac12 \log \det 
  (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
\end{align}
where $\obs$ is constrained to some allowed set of measurements
(e.g.\ pointwise evaluations on $[0,\pi]$). We call this simultaneous
optimization.

For computational reasons, one may prefer to find the best
measurements in a sequential manner. Denote
\begin{equation}\label{eq:def obs_k}
  \obs_k := \sum_{i=1}^k \meas_i\meas^*_i.
\end{equation}
Sequential optimal design proceeds as follows. Find $\meas_1$ by
maximizing $\tar(\obs_1)$. Then, keeping $\meas_1$ fixed --- find
$\meas_2$ as the maximizer of $\tar(\obs_2)$. Then, find $\meas_3$ by
keeping $\meas_1,\meas_2$ fixed and taking $\meas_3$ the maximizer of
$\tar(\obs_3)$. Continue in this fashion until $\obs_m = \obs$ is
found, where $m$ is the number of available measurements. It is
important to notice that this scheme does not require actually
observing data --- in \eqref{eq:objective} data is averaged out.

This paper is mostly focused with the more general simultaneous
optimization case. The sequential optimization case is dealt with in
section \ref{subsec:seq clus}. It is important to note, however that
all conclusions we make on the simultaneous case easily specialize to
the sequential case by considering the posterior as the next
sequential step's prior.

@@

\subsection{Overview}
%%, for the example in section \ref{subsec:example}).
We seek a formulation of the optimal design problem
\eqref{eq:optimization} using Lagrange multipliers. In section
\ref{section:objective} we find the gradient of $\tar$ and record it
in \eqref{eq:tar grad}. In section \ref{subsec:unit length} we relax
the constraints on $\obs$ so that the optimization problem in
\eqref{eq:optimization} is analytically tractable.  We find gradients
for the new constraints and record them in \eqref{eq:grad
  constraints}. Finally, in section \ref{subsec:necessary} we use
these to formulate the optimal design problem as a Lagrange
multipliers problem in equation \eqref{eq:conditions}.

@@

This allow us to arrive at the following:
\begin{corollary}[Eigenvectors of a D-optimal design]\label{cor:same ev}
  $\obs^*\obs$ has the same eigenvectors as $\fwd \prcov \fwd^*$.
  %namely $\{\ev_i\}_{i=1}^{\infty}$.
\end{corollary}
\begin{proof}
In \eqref{eq:mod conditions}, take $\func_j^{*} = \meas_j$ and $C :=
(\fwd \prcov \fwd^*)^{-1}$ and use lemma \ref{lemma:sim diag}.
\end{proof}
Note that all but finitely many of these eigenvectors have zero
eigenvalue, since $\obs^*\obs$ is finite-rank. Denote the
corresponding non-zero eigenvalues of $\obs^*\obs$ by
$\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$. Using
Sylvester's determinant theorem we get:

@@

\begin{figure}%{r}{0.25\textwidth} 
    %\centering
    \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
    \caption{Three graduated lab cylinders, corresponding to three
      eigenvectors. Prior not shown.}
    \label{fig:cylinder}
\end{figure}

The argument above has the important consequence (illustrated in
figure \ref{fig:optimal vs not}) that the eigenvalues in the
eigenvalue problem \eqref{eq:eigenproblem} are all equal! This makes
perfect sense. First, we know the Lagrange multipliers tell us how
much we gain by relaxing the constraints and there is no reason for
any measurement to give us more than any other (otherwise we would put
more weight on the better measurement, as can be understood from the
discussion in the following paragraph).

One can understand optimal designs as follows: Imagine each
eigenvector is a lab cylinder (figure \ref{fig:cylinder}). Each
cylinder $i$ is filled, a-priori, with green liquid
$\sigma^2\lambda_i^{-1}$ volume units (this is the prior precision,
weighted by obsevation noise). An optimal design with (e.g.) four
measurements adds blue liquid of four volume units. It does so by
adding each drop to whatever cylinder has the lowest level of liquid
in it (lowest precision). The result of such a procedure is
illustrated in figure \ref{fig:optimal vs not}. The analysis
conducted above is summarized in theorem \ref{thm:char} below.

\pgfplotstableread{
  Label     prior  optimal  sub-optimal 
  1         0.2    1.8           1.7
  2         0.8    1.2           0.8
  3         2.2    0             0.5
  4         3.5    0             0.0
}\optimalvsnot

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Characterizing an optimal design, compared to a
    sub-optimal design. Both designs are allowed three measurements,
    which means the blue area has accumulated height of $3$ in both
    plots. An optimal design (left) increases precision (reduces
    uncertainty) where it is lowest. A sub-optimal design (right) does
    not. In this example, a measurement is taken partly in the direction of the
    fourth eigenvector, where precision is highest.}
  \label{fig:optimal vs not}
\end{figure}

\pgfplotstableread{
  Label    prior  $o_1$  $o_2$   $o_3$   $o_4$   $o_5$         $o_6$    topper
  1         0.2    0.8    0.5     0.5    0.4    0.33333333  0.3333333  0.001
  2         0.8    0.2    0.5     0.5    0.4    0.33333333  0.3333333  0.001
  3         2.2    0      0       0      0.2    0.33333333  0.3333333  0.001
  4         3.5    0      0       0      0      0           0          0.001
}\clusterization

\pgfplotstableread{
  Label    prior  $o_1$  $o_2$   $o_3$   $o_4$   $o_5$         $o_6$    topper
  1         0.2    0.8    0.7     0.3    0.4    0.2         0.4666666  0.001
  2         0.8    0.2    0.3     0.7    0.4    0.46666666  0.2        0.001
  3         2.2    0      0       0      0.2    0.33333333  0.3333333  0.001
  4         3.5    0      0       0      0      0           0          0.001
}\noclusterization

@@

%% illustrated in figure \ref{fig:clusterization} and

\subsection{Sequential Design}
Recall the sequential optimal design scheme from section
\ref{subsec:seq vs sim} and the definition of $\obs_k$ from
\eqref{eq:def obs_k}. The way clusterization can occur is understood
using results of the previous section, specialized to one measurement
each time. By corollary \ref{cor:ame ev}, the first measurement
$\meas_1$ has to be (parallel to) an eigenvector of
$\fwd\prcov\fwd^*$. By theorem \ref{thm:char}, it will be taken in the direction of the
first eigenvector --- $\meas_1 \parallel \ev_1$. As hinted in
section \ref{subsec:seq vs sim}, the posterior becomes the prior for
the next step. Similar reasoning leads to:
\begin{align*}
  \begin{split}
    \meas_2 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_2 \\
    \meas_2 \parallel \ev_2  &\text{ o.w. }
  \end{split}
\end{align*}
In the first case, we get clusterization. In the second, we consider $\meas_3$:
\begin{align*}
  \begin{split}
    \meas_3 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_3 \text{ (note $\lambda^{-1}_1 < \lambda^{-1}_2$)}  \\
    %\meas_3 \parallel  \ev_2  &\text{ if } \\
    \meas_3 \parallel  \ev_3  &\text{ o.w. }
  \end{split}
\end{align*}
Again, the first case results in clusterization. The sequential design
proceeds in this fashion, where every $\meas_i$ is taken in the
direction of the eigenvector of smallest eigenvalue of the previous
posterior precision operator. Taking enough measurements we will
surely arrive at clusterization, since the prior precision increases
quickly --- in section \ref{subsec:abstract OED} we assumed  $\fwd$
is strongly smoothing, which typpically means eigenvalues of $(\fwd
\prcov\fwd^*)^{-1}$ grow exponentially.

\subsection{Simultaneous Design}
In a simultaneous design scenario, the result is not as clear cut as
in the sequential design case. We still rely on previous results ---
namely corollary \ref{cor:same ev} and theorem \ref{thm:char}. An
illustration of two D-optimal designs scenarios for the same problem
are presented in figure \ref{fig:clusterization} --- the first
exhibits clusterization and second does not. As was discussed earlier
and illustrated in figure \ref{fig:optimal vs not}, an optimal design
makes the precision ``flat'' where it is lowest. Clusterization may
arise when two or more eigenvectors have equal precision. This
happens, for example for eigenvectors $1$ and $2$ when considering
measurements $\meas_i,i=1,2,3$. Taking three measurements, it is
possible that $\meas_1$ is such that it equalizes the precision of
eigenvectors $1$ and $2$. In this case, taking mesurements $\meas_2 =
\meas_3 = \frac{1}{2}(\ev_1 + \ev_2)$ leads to an optimal design that
exhibits clusterization. Similarly, taking $\meas_5 = \meas_6 =
\frac{1}{3}(\ev_1 + \ev_2 + \ev_3)$ also gives rise to an optimal
design that exhibits clusterization. It is important to note that
there are (infinitely) many optimal designs that actually do not
exhibit clusterization --- e.g. the second design presented in figure
\ref{fig:clusterization}. In the following section we postulate what
causes such designs to be observed in realistic experimental design
scenarios.

\begin{figure}
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\clusterization}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector,
      ]

      \addplot [fill=green!80]  table [y=prior, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=blue!60]   table [y=$o_1$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=red!60]    table [y=$o_2$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=black!60]  table [y=$o_3$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=orange!60] table [y=$o_4$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=cyan!60]   table [y=$o_5$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=purple!60] table [y=$o_6$, meta=Label, x expr=\coordindex] {\clusterization};

      \addlegendentry{prior}
      \addlegendentry{$o_1$}
      \addlegendentry{$o_2$}
      \addlegendentry{$o_3$}
      \addlegendentry{$o_4$}
      \addlegendentry{$o_5$}
      \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\noclusterization}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector,
      ]

      \addplot [fill=green!80]  table [y=prior, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=blue!60]   table [y=$o_1$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=red!60]    table [y=$o_2$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=black!60]  table [y=$o_3$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=orange!60] table [y=$o_4$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=cyan!60]   table [y=$o_5$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=purple!60] table [y=$o_6$, meta=Label, x expr=\coordindex] {\noclusterization};

      \addlegendentry{prior}
      \addlegendentry{$o_1$}
      \addlegendentry{$o_2$}
      \addlegendentry{$o_3$}
      \addlegendentry{$o_4$}
      \addlegendentry{$o_5$}
      \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization and non-clusterization in simultaneous
    design. Posterior precision per eigenvector, after taking a
    measurement $\meas_i$. Each $\meas_i$ increases precision
    uniformly across the lowest precision eigenvectors. We see that
    $\meas_2$ and $\meas_3$ are repeated measurements, as well as
    $\meas_5$ and $\meas_6$.}
  \label{fig:clusterization}
\end{figure}

\subsection{Why do Realistic Designs Exhibit Clusterization?}
It is unclear from the previous analysis what makes realistic designs
exhibit clusterization, such as the one observed in figure
\ref{fig:clusterization illustration}. We can only hypothesize.

@@

\subsection{Derivation}
We rely on corollary \ref{cor:same meas} below, which is proved in the
appendix. Denote $\obs = (\meas_1,\dots,\meas_m)^t$ and $\obsm :=
(\meas_1,\dots,\meas_{m-1})^t$.
\begin{restatable*}{corollary}{samemeas}\label{cor:same meas}
  If $\meas_m = \meas_j$ for some $1 \leq j \leq m-1$, then
  \begin{equation*}
    \tar(\obs) - \tar(\obsm) =
    \log \left ( 1 + \frac{\sigma^2
      \langle \fwd \postcovm \fwd^* \obsm^* \Sigmam^{-1} e_j,
      \obsm^* \Sigmam^{-1}e_j \rangle
    }{
      2 - \sigma^2 e_j^t\Sigmam^{-1}e_j 
    }       
    \right ).
  \end{equation*}
\end{restatable*}
Recall from \eqref{eq:Sigma} that $\Sigma(\obs) = \obs \modcov \obs^*
+ \sigma^2I$. From Corollary \ref{cor:same meas} we can deduce that

$$
\lim_{\sigma^2 \to 0} \tar(\obs) -\tar(\obsm) = 0
$$

Hence no increase in the design criterion is achieved by taking a
repeated measurement. This is clearly sub-optimal and $\obsm$ cannot
be an optimal design. Thus, for small measurement error levels, the
clusterization effect will be mitigated by the presence of a non-zero
model error. Since the design criterion is not defined for $\sigma^2 =
0$ and identical measurements, we cannot make a statement regarding
$\sigma^2 = 0$, except in the limiting sense described above.

%% \section{Conclusion and Future Research Directions}\label{section:conclusion}
%% Our relaxed model gives us insight to D-optimal designs. We see that
%% in our setting, uncertainty is not reduced in any direction
%% (eigenvector) before it is the largest uncertainty present. We show
%% how and why repeated measurements can give rise to a D-optimal
%% design. There is more work to be done in understanding exactly what
%% causes such designs to be strictly better than others.

@@

\begin{lemma}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M =
  m$, $m \geq k$. Then we can find $\func_j \in \R^k,j=1,\dots,m$ with
  $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that $AA^t =
  M$.
\end{lemma}
\begin{proof}
  Let us diagonalize $M$, so that $M = U D U^t$ with $D =
  \diag(d_1,\dots,d_k)$ and $U \in \R^{k \times k }$ orthogonal. Let
  $S \in \R^{k \times m}$ with $S_{ii} = \sqrt{d_{i}}$ and zeros
  otherwise. Define $A:= U S V^t$, where $V \in \R^{m \times m}$ is
  orthogonal and will be further restricted later. Then $AA^t = U
  SV^tVS^t U^t = UDU^t$, so $AA^t$ has the required eigenvalues and
  eigenvectors by construction. If we can choose $V$ such that $A$
  also satisfies the unit norm constraints we are done. These
  constraints are, for $j=1,\dots,m$:
  \begin{equation}\label{eq:V constraints}
   1 = [A^tA]_{jj} = [V S^tS V^t]_{jj},
  \end{equation}
  and we can expect to do this since we assumed $\ttr D = m$.

  Define $C = S^tS - I \in \R^{m \times m}$. Note that $\ttr C = 0$ and
  $C$ is diagonal with non-zero entries $d_i-1,i=1,\dots,k$. It suffices
  to find $V$ orthogonal such that $V C V^t$ has zero diagonal. We
  construct such $V$ by sequentially inserting zeros in the diagonal
  and not destroying zeros we already introduced, starting from the
  last diagonal entry and moving to the first. If $c_{mm} \neq 0$ ,
  let $p < m$ such that $c_{pp}c_{mm} < 0$ (such $p$ exists because
  the trace is zero) and let $\theta \in (0,\pi)$. Define a Givens
  rotation $R^{(m)} \in \R^{m \times m}$ by
  \begin{equation*}
    r^{(m)}_{ab} :=
    \begin{cases}
      1 & a = b \neq p \text{ or } a = b \neq m \\
      \cos \theta & a = b = p  \\
     -\sin \theta & a = p, b = m\\
      \cos \theta & a = b = m \\
      \sin \theta & a = m, b = p \\ 
      0 & o.w
    \end{cases}
  \end{equation*}
  Note that conjugating a matrix by $R^{(m)}$ changes only its $m$ and
  $p$ rows and columns. We want to choose $\theta$ such that
  \begin{equation}\label{eq:mm}
    0 = [R^{(m)} C (R^{(m)})^t]_{mm} = \cos^2 \theta c_{mm} + 2\cos \theta \sin
    \theta c_{mp} + \sin^2\theta c_{pp},
  \end{equation}
  and it suffices to choose $\theta$ such that
  \begin{equation*}
    c_{mm} \cot^2 \theta + 2 c_{mp} \cot \theta + c_{pp} = 0.
  \end{equation*}
  This quadratic in $\cot\theta$ has a real solution, since
  $c_{pp}c_{mm} < 0$ by assumption and we can find $\theta \in
  (0,\pi)$ such that \eqref{eq:mm} is satisfied. We continue to find
  $R^{(m-1)}$ that leaves row and column $m$ unchanged and
  continue introducing zeros to the diagonal. The assumption $\ttr D =
  m \Rightarrow \ttr C = 0$ guarantees we can do that. Taking $V:=
  R^{(1)} R^{(2)} \dots R^{(m-1)}R^{(m)}$ completes the proof.
\end{proof}
